DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2023-05-14 manuscript/manuscript.Rmd
add link to suppl figures on zenodo

At the time of writing of this manuscript
<!-- (Jun 22, 2022), -->
(`r format(Sys.time(), "%b %d, %Y")`),
DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata [@mctavish2015phylesystem], which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the `datelife` R package.

@@

In some cases, secondary calibrations might represent the only information available (e.g., groups with a poor fossil record). Recently, @powell2020quantifying showed in a simulation analysis that secondary calibrations are as good as primary calibrations. It seems that using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases described before [@shaul2002playing; @graur2004reading; @sauquet2013practical]. <!--\textcolor{red}{LLSR: What does our cross validation analysis show in relation to  the effect of using secondary calibrations for a dating analysis that does not make any prior assumptions on the nature of the calibrations themselves? the age difference between primary and secondary calibrations?}--> Our cross validation analysis also provides some insight into the application of secondary calibrations. In general, node ages summarized with DateLife and those from the original studies appear positively correlated (Supplementary Figures S2-S20). We also note that nodes closer to the root tend to be slightly younger than ages from the original studies. In contrast, nodes closer to the tips tend to be slightly older when estimated using secondary calibrations than ages from the original studies. The only exception to this trend was observed in @burns2014phylogenetics chronogram, which generally displays much younger node ages when estimated using secondary calibrations than the original study (Supplementary Figs. S1, S5), supporting previous observations [@sauquet2012testing; @schenk2016sec].

@@

Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be found in the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as in the Zenodo stable repositories that host the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), and the software benchmark (<https://doi.org/10.5281/zenodo.7435106>). The corresponding development versions hosted on GitHub can be found at <https://github.com/LunaSare/datelifeMS1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>.

>>>>

At the time of writing of this manuscript
(May 15, 2023),
<!-- (`r format(Sys.time(), "%b %d, %Y")`), -->
DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata [@mctavish2015phylesystem], which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the `datelife` R package.

@@

In some cases, secondary calibrations might represent the only information available for a dating analysis (e.g., groups with a poor fossil record). Recently, @powell2020quantifying showed in a simulation analysis that secondary calibrations are as good as primary calibrations. It seems that using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases described before [@shaul2002playing; @graur2004reading; @sauquet2013practical]. <!--\textcolor{red}{LLSR: What does our cross validation analysis show in relation to  the effect of using secondary calibrations for a dating analysis that does not make any prior assumptions on the nature of the calibrations themselves? the age difference between primary and secondary calibrations?}--> Our cross validation analysis also provides some insight into the application of secondary calibrations. In general, node ages summarized with DateLife and those from the original studies appear positively correlated (Supplementary Figs. S2-S20). We also note that nodes closer to the root tend to be slightly younger than ages from the original studies. In contrast, nodes closer to the tips tend to be slightly older when estimated using secondary calibrations than ages from the original studies. The only exception to this trend was observed in @burns2014phylogenetics chronogram, which generally displays much younger node ages when estimated using secondary calibrations than the original study (Supplementary Figs. S1, S5), supporting previous observations [@sauquet2012testing; @schenk2016sec].

@@

Supplementary Figures can be viewed and dowloaded from their Zenodo repository (<https://doi.org/10.5281/zenodo.6683667>).
Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be downloaded from the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as in the Zenodo stable repositories that host the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), and the software benchmark (<https://doi.org/10.5281/zenodo.7435106>). Development versions corresponding to all of the above are hosted on GitHub, accesible at <https://github.com/LunaSare/datelifeMS1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>.
